\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-04-24 doc/USER_MANUAL/copyright_and_version.tex

%% LyX 1.5.1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[oneside,english,onecolumn]{book}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{letterpaper,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}

% figures
\usepackage[pdftex]{graphicx}

% we are running pdflatex, so convert .eps files to .pdf
\usepackage{epstopdf}

\usepackage{amssymb}
\IfFileExists{url.sty}{\usepackage{url}}
                      {\newcommand{\url}{\texttt}}
\usepackage[authoryear]{natbib}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\newcommand{\noun}[1]{\textsc{#1}}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
{\begin{list}{}{
\setlength{\rightmargin}{\leftmargin}
\setlength{\listparindent}{0pt}% needed for AMS classes
\raggedright
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\normalfont\ttfamily}%
 \item[]}
{\end{list}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\renewcommand{\baselinestretch}{1.5}

% figures
\usepackage[dvips]{epsfig}

\usepackage{wrapfig}

% fonts
\usepackage{times}

% hyperlinks to sections and references
\usepackage[pdftex,bookmarks=true,bookmarksnumbered=true,pdfpagemode=None,pdfstartview=FitH,pdfpagelayout=SinglePage,pdfborder={0 0 0}]{hyperref}

\newcommand{\urlwithparentheses}[1]{(\url{#1})}

% biblio GJI
\bibliographystyle{abbrvnat}

\newcommand{\toall}[1]{\textbf{*** All: #1 ***}}
\newcommand{\tojeroen}[1]{\textbf{*** Jeroen: #1 ***}}
\newcommand{\tobrian}[1]{\textbf{*** Brian: #1 ***}}
\newcommand{\tovala}[1]{\textbf{*** Vala: #1 ***}}
\newcommand{\tovalabrian}[1]{\textbf{*** Vala \& Brian: #1 ***}}
\newcommand{\tovalaqinya}[1]{\textbf{*** Vala \& Qinya: #1 ***}}
\newcommand{\toqinya}[1]{\textbf{*** Qinya: #1 ***}}
\newcommand{\tomin}[1]{\textbf{*** Min: #1 ***}}
\newcommand{\toalessia}[1]{\textbf{*** Alessia: #1 ***}}
\newcommand{\todimitri}[1]{\textbf{*** Dimitri: #1 ***}}

\newcommand{\nexxi}{\mbox{\texttt{NEX\_XI\/}}}
\newcommand{\nexeta}{\mbox{\texttt{NEX\_ETA\/}}}
\newcommand{\nprocxi}{\mbox{\texttt{NPROC\_XI\/}}}
\newcommand{\nproceta}{\mbox{\texttt{NPROC\_ETA\/}}}
\newcommand{\nchunks}{\mbox{\texttt{NCHUNKS\/}}}
\newcommand{\eg}{e.g.,}

% colors for the text
\usepackage[dvipsnames,usenames]{color}
\newcommand{\red}[1]{\textcolor{Red}{#1}}

\usepackage{babel}
\makeatother

\begin{document}
\begin{center}
\thispagestyle{empty}%
\vspace*{-1.8truecm}
\noindent\makebox[\textwidth]{%
\noindent\includegraphics[width=0.83\paperwidth]{figures/specfem_3d-cover.pdf}
}
\end{center}
%
\title{\textbf{SPECFEM3D}\\
\textbf{User Manual}}

\author{$\copyright$ Princeton University (USA) and\\
CNRS / INRIA / University of Pau (France)\\
Version 2.1}

\date{\noindent \today}

\maketitle

\section*{Authors}
The SPECFEM3D package was first developed by Dimitri Komatitsch and Jeroen Tromp at Harvard University and Caltech, USA, starting in 1998, based on earlier work by Dimitri Komatitsch and Jean-Pierre Vilotte at Institut de Physique du Globe (IPGP) in Paris, France from 1995 to 1997.\\

Since then it has been developed and maintained by a development team: in alphabetical order,
Piero Basini,
C\'eline Blitz,
Ebru Bozda\u{g},
Emanuele Casarotti,
Joseph Charles,
Min Chen,
Vala Hj\"orleifsd\'ottir,
Sue Kientz,
Dimitri Komatitsch,
Jes\'us Labarta,
Nicolas Le Goff,
Pieyre Le Loher,
Qinya Liu,
Yang Luo,
Alessia Maggi,
Federica Magnoni,
Roland Martin,
Dennis McRitchie,
Matthias Meschede,
Peter Messmer,
David Mich\'ea,
Tarje Nissen-Meyer,
Daniel Peter,
Max Rietmann,
Brian Savage,
Bernhard Schuberth,
Anne Sieminski,
Leif Strand,
Carl Tape,
Jeroen Tromp,
Zhinan Xie,
Hejun Zhu.\\
%\toall{add full list of developers here and update the AUTHORS file accordingly}

The cover graphic of the manual was created
by Santiago Lombeyda from Caltech's Center for Advanced Computing Research (CACR) \urlwithparentheses{http://www.cacr.caltech.edu}. \\

\newpage{}

\tableofcontents{}

\chapter{Introduction}

The software package SPECFEM3D simulates seismic
wave propagation at the local or regional scale
based upon the spectral-element method (SEM).
The SEM is a continuous Galerkin technique \citep{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11},
which can easily be made discontinous \citep{BeMaPa94,Ch00,KoWoHu02,ChCaVi03,LaWaBe05,Kop06,WiStBuGh10,AcKo11};
it is then close to a particular case of the discontinuous Galerkin technique \citep{ReHi73,Arn82,FaRi99,HuHuRa99,CoKaSh00,GiHeWa02,RiWh03,MoRi05,GrScSc06,AiMoMu06,BeLaPi06,DuKa06,DeSeWh08,PuAmKa09,WiStBuGh10,DeSe10,EtChViGl10}, with optimized efficiency because of its tensorized basis functions \citep{WiStBuGh10,AcKo11}.
In particular, it can accurately handle very distorted mesh elements \citep{OlSe11}.\\

It has very good accuracy and convergence properties \citep{MaPa89,SePr94,DeFiMu02,Coh02,DeSe07,SeOl08}.
The spectral element approach admits spectral rates of convergence and allows exploiting $hp$-convergence schemes.
It is also very well suited to parallel implementation on very large supercomputers \citep{KoTsChTr03,TsKoChTr03,KoLaMi08a,CaKoLaTiMiLeSnTr08,KoViCh10} as well as on clusters of GPU accelerating graphics cards \citep{KoMiEr09,KoErGoMi10,Kom11}. Tensor products inside each element can be optimized to reach very high efficiency \citep{DeFiMu02}, and mesh point and element numbering can be optimized to reduce processor cache misses and improve cache reuse \citep{KoLaMi08a}. The SEM can also handle triangular (in 2D) or tetrahedral (in 3D) elements \citep{WinBoyd96,TaWi00,KoMaTrTaWi01,Coh02,MeViSa06} as well as mixed meshes, although with increased cost and reduced accuracy in these elements, as in the discontinuous Galerkin method.\\

Note that in many geological models in the context of seismic wave propagation studies
(except for instance for fault dynamic rupture studies, in which very high frequencies or supershear rupture need to be modeled near the fault, see e.g. \cite{BeGlCrViPi07,BeGlCrVi09,PuAmKa09,TaCrEtViBeSa10})
a continuous formulation is sufficient because material property contrasts are not drastic and thus
conforming mesh doubling bricks can efficiently handle mesh size variations \citep{KoTr02a,KoLiTrSuStSh04,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09}.\\

For a detailed introduction to the SEM as applied to regional
seismic wave propagation, please consult \citet{PeKoLuMaLeCaLeMaLiBlNiBaTr11,TrKoLi08,KoVi98,KoTr99,ChKoViCaVaFe07} and
in particular \citet{LeKoHuTr09,LeChKoHuTr09,LeChLiKoHuTr08,GoAmTaCaSmSaMaKo09,WiKoScTr04,KoLiTrSuStSh04}.
A detailed theoretical analysis of the dispersion
and stability properties of the SEM is available in \citet{Coh02}, \citet{DeSe07}, \citet{SeOl07} and \citet{SeOl08}.\\

Effects due to lateral variations in compressional-wave speed, shear-wave
speed, density, a 3D crustal model, topography and bathymetry are
included. The package can accommodate
full 21-parameter anisotropy (see~\citet{ChTr07}) as well as lateral
variations in attenuation \citep{SaKoTr10}. Adjoint capabilities and finite-frequency
kernel simulations are included \citep{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11,LiTr06,FiIgBuKe09,ViOp09}.\\

The SEM was originally developed in computational fluid dynamics \citep{Pat84,MaPa89}
and has been successfully adapted to address problems in seismic wave propagation.
Early seismic wave propagation applications of the SEM, utilizing Legendre basis functions and a
perfectly diagonal mass matrix, include \cite{CoJoTo93}, \cite{Kom97},
\cite{FaMaPaQu97}, \cite{CaGa97}, \cite{KoVi98} and \cite{KoTr99},
whereas applications involving Chebyshev basis functions and a nondiagonal mass matrix
include \cite{SePr94}, \cite{PrCaSe94} and \cite{SePrPr95}.\\

All SPECFEM3D software is written in Fortran90 with full portability
in mind, and conforms strictly to the Fortran95 standard. It uses
no obsolete or obsolescent features of Fortran77. The package uses
parallel programming based upon the Message Passing Interface (MPI)
\citep{GrLuSk94,Pac97}.\\

SPECFEM3D won the Gordon Bell award for best performance at the SuperComputing~2003
conference in Phoenix, Arizona (USA) (see \cite{KoTsChTr03}
and \url{www.sc-conference.org/sc2003/nr_finalaward.html}).
It was a finalist again in 2008 for a run at 0.16 petaflops (sustained) on 149,784 processors of the `Jaguar' Cray XT5 system at Oak Ridge National Laboratories (USA) \citep{CaKoLaTiMiLeSnTr08}.
It also won the BULL Joseph Fourier supercomputing award in 2010.\\

This new release of the code (version 2.1) includes support for GPU graphics card acceleration \citep{KoMiEr09,KoGoErMi10,KoErGoMi10,MiKo10,Kom11}.\\

The next release of the code will include support Convolution or Auxiliary Differential Equation Perfectly Matched absorbing Layers
(C-PML or ADE-PML) \citep{KoMa07,MaKoEz08,MaKoGe08,MaKo09,MaKoGeBr10}.
It will also use the PT-SCOTCH parallel library for mesh partitioning.\\

SPECFEM3D includes coupled fluid-solid domains and
adjoint capabilities, which enables one to address seismological inverse problems, but for linear rheologies only so far.
To accommodate visco-plastic or non-linear rheologies, the reader can refer to
the \href{http://geoelse.stru.polimi.it/}{GeoELSE} software package \citep{CaGa97,StPaIg09,ChMoTsBaKrKaStKr10}.

\section{Citation}

If you use SPECFEM3D for your own research, please cite at least one
of the following articles:

\begin{description}

\item [Numerical simulations in general] ~\\
Forward and adjoint simulations are described in detail in \cite{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11,VaCaSaKoVi99,KoMiEr09,KoErGoMi10,KoGoErMi10,ChKoViCaVaFe07,MaKoDi09,KoViCh10,CaKoLaTiMiLeSnTr08,TrKoHjLiZhPeBoMcFrTrHu10,KoRiTr02,KoTr02a,KoTr02b,KoTr99} or \cite{KoVi98}.
Additional aspects dealing with adjoint simulations are described in
\cite{TrTaLi05,LiTr06,TrKoLi08,LiTr08,TrKoHjLiZhPeBoMcFrTrHu10,PeKoLuMaLeCaLeMaLiBlNiBaTr11}.
Domain decomposition is explained in detail in \cite{MaKoBlLe08}, and excellent scaling up to 150,000 processor cores is shown for instance in
\cite{CaKoLaTiMiLeSnTr08,KoLaMi08a,MaKoBlLe08,KoErGoMi10,KoGoErMi10,Kom11},

\item [GPU computing] ~\\
Computing on GPU graphics cards for acoustic or seismic wave propagation applications
is described in detail in \cite{KoMiEr09,KoGoErMi10,KoErGoMi10,MiKo10,Kom11}.

\end{description}

\noindent
If you use this new version 2.1, which has non blocking MPI for much better performance for medium or large runs, please cite at least one of these six articles,
in which results of non blocking MPI runs are presented: \cite{PeKoLuMaLeCaLeMaLiBlNiBaTr11,KoGoErMi10,KoErGoMi10,KoViCh10,Kom11,CaKoLaTiMiLeSnTr08,MaKoBlLe08}.\\

\noindent
If you work on geophysical applications, you may be interested in citing some of these application articles as well, among others:

\begin{description}

\item [Southern California simulations] ~\\
\cite{KoLiTrSuStSh04,KrJiKoTr06a,KrJiKoTr06b}.

If you use the 3D southern California model, please cite
\citet{SuSh03} (Los Angeles model),
\citet{lovelyetal06} (Salton Trough), and
\citet{hauksson2000} (southern California).
The Moho map was determined by
\citet{zhu&kanamori2000}.
The 1D SoCal model was developed by \citet{DrHe90}.

\item [Anisotropy] ~\\
\cite{ChTr07,JiTsKoTr05,ChFaKo04,FaChKo04,RiRiKoTrHe02,TrKo00}.

\item [Attenuation] ~\\
\cite{SaKoTr10,KoTr02a,KoTr99}.

\item [Topography] ~\\
\cite{LeKoHuTr09,LeChKoHuTr09,LeChLiKoHuTr08,GoAmTaCaSmSaMaKo09,WiKoScTr04}.

\end{description}

The corresponding Bib\TeX{} entries may be found
in file \texttt{doc/USER\_MANUAL/bibliography.bib}.

\section{Support}

This material is based upon work supported by the USA National Science
Foundation under Grants No. EAR-0406751 and EAR-0711177, by the French
CNRS, French INRIA Sud-Ouest MAGIQUE-3D, French ANR NUMASIS under
Grant No. ANR-05-CIGC-002, and European FP6 Marie Curie International
Reintegration Grant No. MIRG-CT-2005-017461.
Any opinions, findings, and conclusions or recommendations expressed in this material are
those of the authors and do not necessarily reflect the views of the
USA National Science Foundation, CNRS, INRIA, ANR or the European
Marie Curie program.

\chapter{\label{cha:Getting-Started}Getting Started}

The SPECFEM3D software package comes in a gzipped tar ball. In the
directory in which you want to install the package, type
\begin{lyxcode}
{\small tar~-zxvf~SPECFEM3D\_V2.1.0.tar.gz}{\small \par}
\end{lyxcode}
The directory \texttt{\small SPECFEM3D/} will then contain
the source code.\\

We recommend that you add {\texttt{ulimit -S -s unlimited}} to your {\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize unlimited }} to your {\texttt{.cshrc}} file to suppress any potential limit to the size of the Unix stack.\\

Then, to configure the software for your system, run the
\texttt{configure} shell script. This script will attempt to guess
the appropriate configuration values for your system. However, at
a minimum, it is recommended that you explicitly specify the appropriate
command names for your Fortran90 compiler and MPI package:

\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90
\end{lyxcode}

Before running the \texttt{configure} script, you should probably edit file \texttt{flags.guess} to make sure that it contains the best compiler options for your system. Known issues or things to check are:

\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine.
\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave} or \texttt{-qnosave} for your machine.
\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
In addition, the \texttt{clock\_gettime} routine, which is used by the \texttt{SCOTCH} library that we use, does not exist in Mac OS.
You will need to replace it with \texttt{clock\_get\_time} if you want to use \texttt{SCOTCH}.
\end{description}

When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc} compilers, we suggest running the \texttt{configure} script
with the following options:\\

\noindent\texttt{./configure FC=xlf90\_r MPIFC=mpif90 CC=xlc\_r CFLAGS="-O3 -q64" FCFLAGS="-O3 -q64" -with-scotch-dir=...}.\\

On SGI systems, \texttt{flags.guess} automatically informs \texttt{configure}
to insert `\texttt{`TRAP\_FPE=OFF}'' into the generated \texttt{Makefile}
in order to turn underflow trapping off.\\

If you run very large meshes on a relatively small number
of processors, the memory size needed on each processor might become
greater than 2 gigabytes, which is the upper limit for 32-bit addressing.
In this case, on some compilers you may need to add \texttt{``-mcmodel=medium}'' or \texttt{``-mcmodel=medium -shared-intel}''
to the configure options of CFLAGS, FCFLAGS and LDFLAGS otherwise the compiler will display an error
message (for instance \texttt{``relocation truncated to fit: R\_X86\_64\_PC32 against .bss''} or something similar);
on an IBM machine with the \texttt{xlf} and \texttt{xlc} compilers, using \texttt{-q64} is usually sufficient.\\

The SPECFEM3D software package relies on the SCOTCH library to partition meshes created with CUBIT.
Note that we use CUBIT to create meshes of hexahedra, but other packages can be used as well, for instance GiD from \url{http://gid.cimne.upc.es} or Gmsh from \url{http://geuz.org/gmsh} \citep{GeRe09}. Even mesh creation packages that generate tetrahedra, for instance TetGen from \url{http://tetgen.berlios.de}, can be used because each tetrahedron can then easily be decomposed into four hexahedra as shown in the picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif}; while this approach does not generate hexahedra of optimal quality, it can ease mesh creation in some situations and it has been shown that the spectral-element method can very accurately handle distorted mesh elements \citep{OlSe11}.

The SCOTCH library \citep{PeRo96}
provides efficent static mapping, graph and mesh partitioning routines. SCOTCH is a free software package developed by
Fran\c {c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}. It is more recent than METIS, actively maintained and performs better in many cases.
A recent version of its source code is provided in directory \texttt{src/decompose\_mesh\_SCOTCH/scotch\_5.1.10b}.
In case no SCOTCH libraries can be found on the system, the configuration will bundle this version for compilation.
The path to an existing SCOTCH installation can to be set explicitly with the option \texttt{-{}-with-scotch-dir}. Just as an example:
\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90~-{}-with-scotch-dir=/opt/scotch
\end{lyxcode}

If you use the Intel ifort compiler to compile the code, we recommend that you use the Intel icc C compiler to compile Scotch, i.e., use:
\begin{lyxcode}
./configure~CC=icc~FC=ifort~MPIFC=mpif90
\end{lyxcode}

To compile a serial version of the code for small meshes that fits
on one compute node and can therefore be run serially, run \texttt{configure}
with the \texttt{-{}-without-mpi} option to suppress all calls to
MPI.

A summary of the most important configuration variables follows.

\begin{description}
\item [{\texttt{F90}}] Path to the Fortran90 compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran90.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flag for non-critical subroutines.
\item [{\texttt{FLAGS\_NO\_CHECK}}] Compiler flag for creating fast, production-run
code for critical subroutines.
\end{description}
The configuration script automatically creates for each executable a corresponding
\texttt{Makefile} in the \texttt{src/} subdirectoy.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{constants.h} file depending on your
system:

\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database for the solver on the
local disks of the compute nodes. Some systems have no local disks,
e.g., BlueGene or the Earth Simulator, and other systems have a fast
parallel file system, in which case this flag should be set to \texttt{.true.}.
Note that this flag is not used by the database generator or the solver; it is
only used for some of the post-processing.
\end{description}
The package can run either
in single or in double precision mode. The default is single precision
because for almost all calculations performed using the spectral-element method
using single precision is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half as much memory.
Select your preference by selecting the appropriate setting in the \texttt{constants.h} file:

\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
In the \texttt{precision.h} file:

\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
On many current processors (e.g., Intel, AMD, IBM Power), single precision calculations
are significantly faster; the difference can typically be 10\%
to 25\%. It is therefore better to use single precision.
What you can do once for the physical problem you want to study is run the same calculation in single precision
and in double precision on your system and compare the seismograms.
If they are identical (and in most cases they will), you can select single precision for your future runs.\\

\section{Adding OpenMP support in addition to MPI}

OpenMP support can be enabled in addition to MPI. However, in many cases performance will not improve
because our pure MPI implementation is already heavily optimized and thus the resulting code will in fact
be slightly slower. A possible exception could be IBM BlueGene-type architectures.

To enable OpenMP, uncomment the OpenMP compiler option in two lines in file \texttt{src/specfem3D/Makefile.in}
(before running \texttt{configure}) and also uncomment the \texttt{\#define USE\_OPENMP} statement in file\\
\texttt{src/specfem3D/specfem3D.F90}.

The DO-loop using OpenMP threads has a SCHEDULE property. The OMP\_SCHEDULE
environment variable can set the scheduling policy of that DO-loop.
Tests performed by Marcin Zielinski at SARA (The Netherlands) showed that often
the best scheduling policy is DYNAMIC with the size of the chunk equal to the number of
OpenMP threads, but most preferably being twice as the number of
OpenMP threads (thus chunk size = 8 for 4 OpenMP threads etc).
If OMP\_SCHEDULE is not set or is empty, the DO-loop will assume generic
scheduling policy, which will slow down the job quite a bit.

\section{Visualizing the subroutine calling tree of the source code}

Packages such as \texttt{doxywizard} can be used to visualize the subroutine calling tree of the source code.
\texttt{Doxywizard} is a GUI front-end for configuring and running \texttt{doxygen}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{\label{cha:Mesh-Generation}Mesh Generation}

The first step in running a spectral-element simulation consists of constructing a high-quality mesh for the region under consideration. We provide two possibilities to do so: (1) relying on the external, hexahedral mesher CUBIT, or (2) using the provided, internal mesher \texttt{xmeshfem3D}. In the following, we explain these two approaches.

\section{\label{cha:Running-the-Mesher-CUBIT}Meshing with \texttt{CUBIT}}

CUBIT is a meshing tool suite for the creation of finite-element meshes for arbitrarily shaped models. It has been developed and maintained at Sandia National Laboratories and can be purchased for a small academic institutional fee at \url{http://cubit.sandia.gov}.
Our experience showed that using CUBIT greatly facilitates and speeds up the generation and preparation of hexahedral, conforming meshes for a variety of geophysical models with increasing complexity.

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.6\textwidth]{figures/mount-cubit.jpg}
\par\end{centering}

\caption{Example of the graphical user interface of CUBIT. The hexahedral mesh shown in the main display consists of a hexahedral discretization of a single volume with topography.}

\label{fig:mount.cubit}
\end{figure}

The basic steps in creating a load-balanced, partitioned mesh with CUBIT are:
\begin{description}
\item [1.] setting up a hexahedral mesh with CUBIT,
\item [2.] exporting the CUBIT mesh into a SPECFEM3D file format and
\item [3.] partitioning the SPECFEM3D mesh files for a chosen number of cores.
\end{description}

Examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. We strongly encourage you to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \urlwithparentheses{cig-seismo@geodynamics.org}.

\subsection{Creating the Mesh with CUBIT}

For the installation and handling of the CUBIT meshing tool suite, please refer to the CUBIT user manual and documentation.
In order to give you a basic understanding of how to use CUBIT for our purposes, examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}:
\begin{description}
\item [{\texttt{homogeneous\_halfspace}}] Creates a single block model and assigns elastic material parameters.
\item [{\texttt{layered\_halfspace}}]  Combines two different, elastic material volumes and creates a refinement layer between the two. This example can be compared for validation against the solutions provided in subdirectory ~\\
\texttt{VALIDATION\_3D\_SEM\_SIMPLER\_LAYER\_SOURCE\_DEPTH/}.
\item [{\texttt{waterlayered\_halfspace}}] Combines an acoustic and elastic material volume as in a schematic marine survey example.
\item [{\texttt{tomographic\_model}}] Creates a single block model whose material properties will have to be read in from a tomographic model file during the databases creation by \texttt{xgenerate\_databases}.
\end{description}

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.45\textwidth]{figures/example-homogeneous.jpg}
\includegraphics[width=.45\textwidth]{figures/example-2layers.jpg} \\
\includegraphics[width=.45\textwidth]{figures/example-water.jpg}
\includegraphics[width=.45\textwidth]{figures/example-tomo.jpg}
\par\end{centering}

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace (top-left), layered halfspace (top-right), water layered halfspace (bottom-left) and tomographic model (bottom-right).}

\label{fig:examples.cubit}
\end{figure}

In each example subdirectory you will find a \texttt{README} file, which explains in a step-by-step tutorial the workflow for the example.
Please feel free to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \urlwithparentheses{cig-seismo@geodynamics.org}.

\subsection{Exporting the Mesh with \texttt{cubit2specfem3d.py} }
\label{subsec:Exporting-the-Mesh}

Once the geometric model volumes in CUBIT are meshed, you prepare the model for exportation with the definition of material blocks and boundary surfaces. Thus, prior to exporting the mesh, you need to define blocks specifying the materials and absorbing boundaries in CUBIT.
This process could be done automatically using the script \texttt{boundary\_definition.py} if the mesh meets some conditions or manually, following the block convention:

\begin{description}
\item [material\_name] Each material should have a specific block defined by a unique name. The name convention of the material is to start with either 'elastic' or 'acoustic'. It must be then followed by a unique identifier, e.g. 'elastic 1', 'elastic 2', etc.
The additional attributes to the block define the material description.

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-wave speed of the material (given in m/s).
\item [Vs] S-wave speed of the material (given in m/s).
\item [rho] density of the material (given in kg/m$^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note that your Vp- and Vs-speeds are given for a reference frequency. To change this reference frequency, you change the value of
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

For an acoustic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-wave speed of the material (given in m/s).
\item [0] S-wave speed of the material is ignored.
\item [rho] density of the material (given in kg/m$^3$).
\end{description}

\item [face\_topo] Block definition for the surface which defines the free surface (which can have topography). The name of this block must be 'face\_topo', the block has to be defined using all the surfaces which constitute the complete free surface of the model.

\item [face\_abs\_xmin] Block definition for the faces on the absorbing boundaries, one block for each surface with x=Xmin.
\item [face\_abs\_xmax] Block definition for the faces on the absorbing boundaries, one block for each surface with x=Xmax.
\item [face\_abs\_ymin] Block definition for the faces on the absorbing boundaries, one block for each surface with y=Ymin.
\item [face\_abs\_ymax] Block definition for the faces on the absorbing boundaries, one block for each surface with y=Ymax.
\item [face\_abs\_bottom] Block definition for the faces on the absorbing boundaries, one block for each surface with z=bottom.
\end{description}

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=.45\textwidth]{figures/mount-surface.jpg}
\includegraphics[width=.45\textwidth]{figures/mount-abs.jpg}
\par\end{centering}

\caption{Example of the block definitions for the free surface 'face\_topo' (left) and the absorbing boundaries, defined in a single block 'face\_abs\_xmin' (right) in CUBIT.}

\label{fig:mount.abs}
\end{figure}

Optionally, instead of specifying for each surface at the boundaries a single block like mentioned above, you can also specify a single block for all boundary surfaces and name it as one of the absorbing blocks above, e.g. 'face\_abs\_xmin'.

After the block definitions are done, you export the mesh using the script \texttt{cubit2specfem3d.py} provided in each of the example directories. If the export was successful, you should find the following files in a subdirectory \texttt{MESH/}:

\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag}
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-wave speed in $m \, s^{-1}$, \texttt{vs}  the S-wave speed in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.

%%magnoni
For tomographic velocity models, please read Chapter \ref{cha:-Changing-the} and Section \ref{sec:Using-tomographic} `Using external tomographic Earth models' for further details.
%
%For a tomographic model, the material definition format is:
%\begin{lyxcode}
%\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}
%\end{lyxcode}
%where
%\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file \texttt{model\_tomography.f90} located in the
%\texttt{src/generate\_databases/} directory.
%
\item [materials\_file] Contains the material associations for each element.
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity. You should put both the surface of acoustic regions and of elastic regions in that file;
that is, list all the element faces that constitute the surface of the model in that file.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

These mesh files are needed as input files for the partitioner  \texttt{xdecompose\_mesh\_SCOTCH}
to load-balance the mesh. Please see the next section for further details.

\subsubsection*{Checking the mesh quality}
The quality of the mesh may be inspected more precisely based upon the serial code
in the file  \texttt{check\_mesh\_quality\_}~\\
\texttt{CUBIT\_Abaqus.f90} located in the directory \texttt{src/check\_mesh\_quality\_CUBIT\_Abaqus/}.
Running this code is optional because no information needed by the
solver is generated.

>>>>

\addcontentsline{toc}{chapter}{Copyright}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Copyright}

Main historical authors: Dimitri Komatitsch and Jeroen Tromp

Princeton University, USA, and CNRS / University of Marseille, France

$\copyright$ Princeton University and CNRS / University of Marseille, July 2012
